---
title: Practical Machine Learning Project - Quantified Self Movement Data
author: "Pratik Chhapolika"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## ANALYSIS
In the aforementioned study, six participants participated in a dumbell lifting exercise five different ways. The five ways, as described in the study, were exactly according to the specification (Class A), throwing the elbows to the front (Class B), lifting the dumbbell only halfway (Class C), lowering the dumbbell only halfway (Class D) and throwing the hips to the front (Class E). Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.

By processing data gathered from accelerometers on the belt, forearm, arm, and dumbell of the 
participants in a machine learning algorithm, the question is can the appropriate activity quality 
(class A-E) be predicted?

## Data Preprocessing/Cleaning

```{r,echo=TRUE,warning=FALSE,cache=FALSE}

# Loading required R package. If not installed,please use install.packages("Package_Name") to install package first.
# install.packages("caret")
# install.packages("rpart")
# install.packages("rpart.plot")
# install.packages("corrplot")
# install.packages("randomForest")
# install.packages("e1071")
```

```{r,echo=TRUE,warning=FALSE,cache=FALSE,message=FALSE}
# Loading Required Package

library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
library(e1071)
```

```{r,echo=TRUE}

# Load Training and Testing Dataset in R environment
training<-read.csv("pml-training.csv")
test<-read.csv("pml-testing.csv")
dim(training)
# 19622 rows and 160 columns before cleaning data

training <- training[, colSums(is.na(training)) == 0]
test<- test[, colSums(is.na(test)) == 0]
dim(training)

# 19622 rows and 93 columns after cleaning data. Deleting Columns
# which has NA values.

# Extracting the Outcome Variable and storing that in classe Variavle
classe <- training$classe

# Next, we get rid of some columns that do not contribute much
# to the accelerometer measurements.

trainRemove <- grepl("^X|timestamp|window", names(training))
# These are the columns to be removed

names(training)[trainRemove]

training <- training[, !trainRemove]
training <- training[, sapply(training, is.numeric)]
training$classe <- classe
dim(training)
# 19622 rows and 54 columns After cleaning data.

# Similarly for Test dataset
testRemove <- grepl("^X|timestamp|window", names(test))
test<- test[, !testRemove]
test<- test[, sapply(test, is.numeric)]
```
### Slice the data

```{r,echo=TRUE}

# Then, we can split the cleaned training set into a pure training data
# set (70%) and a validation data set (30%). We will use the validation
# data set to conduct cross validation in future steps.

set.seed(1234567) # For reproducibile purpose
inTrain <- createDataPartition(training$classe, p=0.70, list=F)
trainData <- training[inTrain, ]
Cross_Val_Data <- training[-inTrain, ]

```

## Data Modeling
We fit a predictive model for activity recognition using **Random Forest** algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general. We will use **5-fold cross validation** when applying the algorithm. 

```{r, cache = T}

controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainData, method="rf", trControl=controlRf, ntree=250)
modelRf


predictRf <- predict(modelRf, Cross_Val_Data)
confusionMatrix(Cross_Val_Data$classe, predictRf)

accuracy <- postResample(predictRf, Cross_Val_Data$classe)
accuracy

# Accuracy     Kappa
# 0.9998301   0.9997851

oose <- 1 - as.numeric(confusionMatrix(Cross_Val_Data$classe, predictRf)$overall[1])
oose
# So, the estimated accuracy of the model is 99.9

result <- predict(modelRf, test[, -length(names(test))])
result

```
## Appendix: Figures

1. Correlation Matrix Visualization  
```{r, cache = T}
corrPlot <- cor(trainData[, -length(names(trainData))])
corrplot(corrPlot, method="color")
```

2. Decision Tree Visualization
```{r, cache = T}
treeModel <- rpart(classe ~ ., data=trainData, method="class")
prp(treeModel) # fast plot

```


